---
title: "PML exercise"
author: "Gianluca Emireni"
date: "February 9th 2015"
output: html_document
---

---
# Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. 
# These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, 
# to find patterns in their behavior, or because they are tech geeks. 
# One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, 
# your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. 
# They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 
# More information is available from the website here: http://groupware.les.inf.puc-rio.br/har 
# (see the section on the Weight Lifting Exercise Dataset). 
--- 
Loading libraries
```{r, eval=T, }
library(randomForest)
library(caret)
```

Reading datasets
```{r}
train <- read.csv("/Users/gianluca/Documents/PML/pml-training.csv", sep=",", quote="\"", head=T)
test <- read.csv("/Users/gianluca/Documents/PML/pml-testing.csv", sep=",", quote="\"", head=T)
```


Adjust timestamps and remove record counter, not functional for classification
```{r}
train$cvtd_timestamp <- as.numeric(strptime(as.character(train$cvtd_timestamp), format = "%d/%m/%Y %H:%M"))
train <- train[,-1]
```


Some variables are read as factors, I transform them in numeric
```{r, echo=FALSE}
for(i in 1:ncol(train))
{
  col <- colnames(train)[i]
  print(c(col, class(col)))
  if( is.factor(train[,col]) & !(col %in% c("classe","user_name","new_window")) ) {
    train[,col] <- as.numeric(as.character(train[,col]))
  }
}
```

Variables with no valid values (not NA, NULL, NaN, etc.) are excluded from the sets
```{r, echo=FALSE}
invalidRegressors <- c()
for(i in 1:ncol(train))
{
  col <- colnames(train)[i]
  if(all(is.na(train[,col]))) {
    invalidRegressors <- c(invalidRegressors, i)
  }
}
```

Excluding constant variables and variables with near zero variance
```{r, echo=FALSE}
nzv <- nearZeroVar(train)
train <- train[,-nzv]
test <- test[,-nzv]
```

# highly correlated variables
# findCorrelation(cor(train), cutoff = .75)

Filling NA values (only numeric regressors, excluding factors)
```{r}
factor_indexes <- which(colnames(train) %in% c("classe","user_name"))
# 1 5 150
preObj <- preProcess(train[,-factor_indexes], na.remove=T, method = c("center", "scale", "knnImpute"))
# standardizing (BoxCox ?)
stdTrain <- predict(preObj, train[,-factor_indexes])
```

Re-attaching factors
```{r}
stdTrain$user_name <- train$user_name
stdTrain$classe <- train$classe
```


Transforming factors to dummies (excluding the response variable *classe*)
```{r}
train2 <- data.frame(model.matrix(classe ~ . -1, data=stdTrain))
train2$classe <- train$classe
```


# Random Forest Cross-Valdidation for feature selection
rfcv(trainx=subset(train2, select = -classe), trainy=train2$classe, na.action = na.exclude)
rf <- randomForest(x=subset(train, select = -classe), y=train$classe, data=train, xtest = subset(test, select = -classe), ytest = test$classe, ntree = 100)



levels(train$new_window)[1:10]
cor(subset(train, select = -classe))